---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{dplyr}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{rbenvo}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```


```{r libraries,message=F,results='hide'}
library(rsstap)
library(rbenvo)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
data("example_benvo")
```

# Introduction and Motivation

The [`rstap` package](https://biostatistics4socialimpact.github.io/rstap) introduced Spatial Temporal Aggregated Predictor or STAP models, as a method by which the effect of [Built Environment](https://en.wikipedia.org/wiki/Built_environment) Features (BEFs) measured as a point pattern could be incorporated into pre-existing regression methods. The `rsstap` package builds on this framework by modifying the way in which the spatial temporal exposure effect is estimated.

# Model Specification

Restricting our attention to only spatial exposure for the following illustrative example, the spatial aggregated predictor both `rstap` and `rsstap` fit models of the following form:

$$
E[g(\mu_i)] = Z_i^{T}\delta + f(\mathcal{D}_i) \tag{1}\\
$$
While link function $g(\cdot)$, univariate mean $\mu_i$ and covariates $Z_i$ for subject $i=1,...,n$ are identified from a typical regression model, the function $f(\mathcal{D})$ novelly represents the $i$th subject's cumulative exposure across space. In `rstap` This was accomplished by incorporating distances, $d$ through a spatial exposure function $f(\mathcal{D}_i) = \sum_{d \in \mathcal{D}_i} \mathcal{K}_s(d,\theta)$ - typically some exponential decay function, like $\exp(-\frac{d}{\theta})$ or $\exp(-(\frac{d}{\theta})^{\eta})$. Unfortunately, estimating this model through typical methods (e.g. MCMC) requires numerous evaluations of the kernel function summed across the, again typically numerous, distances in the set $\mathcal{D}_i$, of subject-BEF distances.

To remedy this situation, the `rsstap` package adopts a regression spline basis representation, $f(d) = \sum_{j=0}^{J} \beta_j\phi_j(d)$, of the spatial exposure effect, removing the kernel and allowing for estimation on a sum of the function of the distances:

$$
E[g(\mu_i)] = Z_i^{T}\delta +   \sum_{j=0}^{J}\sum_{d \in \mathcal{D}_i}\beta_j\phi_j(d) \tag{2}
$$

Since the sum across distances, $\sum_{d\in \mathcal{D}_i} \phi_j(d)$ can be done prior to model fitting, standard methods (e.g. a bayesian equivalent to `mgcv::gam`) can be used, dramatically decreasing computational complexity. 

## Illustration



```{r auxiliary,echo=F}
truth <- geom_line(aes(x=x,y=y),
            ## True generating function
            data=tibble(x=seq(from=0,to=1,by=0.01),
                        y=3*pweibull(seq(from=0,to=1,by=0.01),
                                     shape = 5,
                                     scale = .6,
                                     lower.tail = F)),
            color='red')
```

```{r model_fit,message=F,results='hide'}
fit <- sstap_lm(BMI ~ sex + sap(FFR),
                example_benvo)
```

We'll examine the model fit by first looking at a print out summary of the model followed by the posterior predictive checks - a basic model check tool.

```{r printfit}
fit
```


```{r ppcs}
ppc(fit)
```

Next we'll take a look at the spatial effect of one single FFR on the expected BMI of subjects as a function of space. We'll overlay the true function (since this is simulated data) in red on top.


```{r plot_results}
plot(fit) + truth
```


While the curves won't always be so well defined, this more or less defines the basics of fitting stap models using the spline implimentation via rsstap. The next section will go through the basics of spatial-temporal estimation *without* adjusting for within subject correlation.


## Longitudinal Spatio-Temporal example

In this quick example, we'll briefly run through the same function calls as before, now using a longitudinal simulated dataset that contains both spatial and temporal data. Note that while a longitudinal model *should* take into account correlation across a given subject's measurements, we'll stick to the simpler model now for exposition (see the longitudinal vignette for how to account for these terms).  Let's first look at the `longitudinal_HFS` data that comes with the `rsstap package`, renamed here as `bdf2` for brevity.

```{r long_data,echo=F}
data("longitudinal_HFS")
bdf2 <- longitudinal_HFS
```

```{r bdf_print}
bdf2
```

Looks like this data contains 300 subjects with both distances and times measured for the subjects. In order to fit the same kind of model as before, but now taking into account both space and time, we'll use the `stap` designation in the model formula. This is equivalent to using the `t2` formula from `mgcv`.

```{r,results='hide'}
fit <- sstap_lm(BMI ~ sex + stap(HFS), benvo = bdf2)
```

We'll examine the posterior predictive checks first to verify that indeed the model can recover the true simulated parameters.

```{r ppc_plot}
ppc(fit)
```


```{r,echo=F}
truth <- geom_contour(aes(x=x,y=y),
            ## True generating function
            data=tibble(x=seq(from=0,to=1,by=0.01),
                        y=-1*pweibull(seq(from=0,to=1,by=0.01),
                                     shape = 5,
                                     scale = .6,
                                     lower.tail = F)),
            color='red')
```

Then, onto the spatio-temporal effects. Calling the plot method on this fitted object will produce a contour plot. `rsstap` also provides a `plot_xsection` which plots the cross section of the surface at a fixed value and a `plot3D` function via the `rplotly` package that can plot the surface in 3 dimensions.

```{r,messages=F,warning=F}
plot(fit)
```

```{r xsection}
plot_xsection(fit,fixed_val=5)
```


```{r 3dplot,warning=FALSE}
plot3D(fit)
```

## Future work

This concludes the introductory vignette. With built environment - subject data, the `rsstap` package can be used to estimate how nearby built environment features may effect the outcome of interest via the STAP modeling framework.

Future vignettes will elaborate on the spline implementation of spatial temporal aggregated predictors in a longitudinal context, wherein subject specific and between/within effects are estimated, network effects of BEFs and spatio-temporal interaction effects.

